---
title: "Generalized S-X2 and conditional infit"
subtitle: "Calibrating S-X2 for Rasch models, and what it adds to infit"
author: "Magnus Johansson"
date: 2026-10-05
date-modified: 2026-10-05
---

```{r common}
#| include: false
source("_common.R")
suppressMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
```

::: {.callout-note appearance="simple" icon=false}
## About this chapter

**Status:** complete<br>
**Simulated:** 2026-10-04 and 2026-10-05 with easyRasch2 1.3.1.9002 and mirt
1.46.1. Rendered from the cached results with `r computed_with(c("easyRasch2", "mirt"))`.<br>
**Cite as:** Johansson, M. (2026). Generalized S-X2 and conditional infit. In
*Rasch Measurement Explorations*`r modified_note()`.
<https://pgmj.github.io/rasch_book/chapters/sx2-infit.html>
:::

::: {.callout-tip appearance="simple" icon=false}
## Summary

S-X2 as computed by `mirt::itemfit()` rejects 6 to 7.5 percent of items at a
nominal 5 under a true Rasch model, because mirt subtracts the estimated latent
variance from every item's degrees of freedom. Restoring that degree of freedom
makes the per-item test nominal, but a smaller excess in the far tail keeps the
family-wise rate at 6 to 9 percent for 20 dichotomous items. A parametric
bootstrap of S-X2 brings it close to 5 percent, with a slight excess at small n.
Held to a family-wise error rate, conditional infit with Westfall-Young detects
at least as much misfit as every S-X2 variant, and clearly more when misfit is
subtle. S-X2's advantage is elsewhere: when several items misfit at n = 1000 or
more, it rarely flags the fitting items that infit flags as well, unless misfit
is strong and n reaches 2000. That resistance belongs to the S-X2 statistic,
not to the bootstrap, whose role is to hold the error rate under the null.
Asymptotic S-X2 with the corrected df and Holm gives nearly the same result at
no extra cost, at a family-wise error rate of 6 to 9 percent. S-X2 is a second
opinion at large n, not a replacement for infit.
:::

## Purpose {#sec-sx2-purpose}

Conditional infit with simulation-based cutoffs detects misfit well, but when
several items misfit and the sample is large it also flags many items that fit,
a cascade documented by @johansson2025. The generalized S-X2 index
[@orlando2000; @kang2008] is a natural candidate for a complement. It compares
observed and expected proportions within sum-score groups, and @kang2008 report
nominal Type I error from n = 500 to n = 5000. Two features of that literature
leave the question open. Power was evaluated only where every item misfit, and
the follow-up on the graded response model found power "much lower when the
source of misfit was multidimensionality" [@kang2011], which is the misfit that
matters most in applied Rasch analysis.

This chapter first checks whether S-X2 holds its nominal level under a true
Rasch model (@sec-sx2-null), finds that it does not and why, and then compares
calibrated versions of S-X2 with conditional infit on the same datasets, with
every method held to a family-wise error rate (@sec-sx2-comparison).

Every replication in both studies generates new data. Designs that draw every
replication from one large simulated dataset leave each sample unbiased but
make the replications dependent, so their Monte Carlo SEs are too small and
differences between sample sizes can be noise from that one dataset.

## S-X2 under a true Rasch model {#sec-sx2-null}

### The degrees of freedom {#sec-sx2-df}

mirt computes the degrees of freedom of S-X2 for each item as the number of
score groups left after collapsing sparse cells, minus the item's own
parameters, minus every estimated latent parameter. For `itemtype = "Rasch"`
the latent variance is estimated, so each item loses one degree of freedom that
the definition of @orlando2000 does not remove. The variance acts as a slope
shared by all items, and charging it to every item counts it once per item. The
deduction is deliberate: mirt's NEWS lists it under version 1.20.1 as a df
adjustment "for models where the latent trait hyper-parameters have been
estimated".

If S-X2 follows a chi-square distribution on df + 1 and is referred to df, the
rejection rate at .05 is 7.5 percent at df = 10 and 6.9 percent at df = 17, the
range of df that 20 dichotomous items give between n = 150 and 2000.

### Design {#sec-sx2-null-design}

- **Item sets.** 20 dichotomous items evenly spaced on [-2, 2]. 20 dichotomous
  items at clustered locations from a single `runif(20, -2, 2)` draw (seed
  20250727), to check that the result does not depend on even spacing. Nine
  items with four categories, locations evenly spaced on [-1.5, 1.5], thresholds
  at location -1.2, 0 and +1.2.
- **Persons.** $\theta \sim N(0, 1.5^2)$.
- **Sample sizes.** 150, 500 and 2000, with 2000 replications per cell. The
  Monte Carlo SE of a 5 percent rate is about 0.5 points.

Four variants of S-X2, from at most two fits per replication:

| Variant | Fit | Reference |
|---|---|---|
| V0 | Rasch, variance estimated | chi-square on mirt's df |
| V1 | same as V0 | chi-square on df + 1 |
| V2 | same as V0, `mincell = 5` | mirt's df, and df + 1 (V2c) |
| V3 | Rasch, variance fixed at the true 2.25 | mirt's df, which then has no latent deduction |

V1 tests the df. V3 removes the deduction and the estimation error in the
variance together, so comparing V1 with V3 separates the two. V2 tests stricter
collapsing of sparse score groups. The simulation is
`scripts/sx2/sx2_null_diagnostic.R`. No replication was discarded.

```{r null-tidy}
dg <- readRDS("results/sx2/null_diagnostic.rds")
dg_ok <- !vapply(dg$res, is.null, logical(1))
dg_long <- do.call(rbind, lapply(which(dg_ok), function(i) {
  g <- dg$grid[i, ]; m <- dg$res[[i]]
  data.frame(set = g$set, n = g$n, rep = g$rep, item = seq_len(nrow(m)), m)
})) |>
  mutate(p_v1  = pchisq(S_v0, df_v0 + 1, lower.tail = FALSE),
         p_v2c = pchisq(S_v2, df_v2 + 1, lower.tail = FALSE),
         set = factor(set, levels = c("dich_even", "dich_orig", "poly"),
                      labels = c("20 dichotomous, even", "20 dichotomous, clustered",
                                 "9 items, 4 categories")))

dg_p <- dg_long |>
  select(set, n, rep, item, V0 = p_v0, V1 = p_v1, V2 = p_v2, V2c = p_v2c,
         V3 = p_v3) |>
  pivot_longer(V0:V3, names_to = "variant", values_to = "p")

v_col <- c(V0 = "#4a3aa7", V1 = "#eb6834", V3 = "#eb6834")
v_lty <- c(V0 = "solid", V1 = "solid", V3 = "dashed")
v_shp <- c(V0 = 18, V1 = 17, V3 = 2)
v_lab <- c(V0 = "mirt df (V0)", V1 = "df + 1 (V1)", V3 = "variance fixed (V3)")

variant_plot <- function(d, ylab, band = NULL) {
  p <- ggplot(d, aes(factor(n), v, colour = variant, linetype = variant,
                     shape = variant, group = variant))
  if (!is.null(band)) {
    p <- p + annotate("rect", xmin = -Inf, xmax = Inf, ymin = 5 - band,
                      ymax = 5 + band, fill = "grey85", alpha = 0.6)
  }
  p +
    geom_hline(yintercept = 5, linetype = "dashed", colour = "grey40") +
    geom_line(linewidth = 0.6) +
    geom_point(size = 2.2) +
    scale_colour_manual(values = v_col, labels = v_lab, name = NULL) +
    scale_linetype_manual(values = v_lty, labels = v_lab, name = NULL) +
    scale_shape_manual(values = v_shp, labels = v_lab, name = NULL) +
    facet_wrap(~ set) +
    labs(x = "Sample size", y = ylab) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}
```

### Per-item rejection {#sec-sx2-null-peritem}

```{r}
#| label: fig-sx2-null-peritem
#| fig-width: 8
#| fig-height: 3.8
#| fig-cap: "Per-item rejection rate of S-X2 at .05 under a true Rasch model,
#|   2000 datasets per sample size. The dashed horizontal line is the nominal 5
#|   percent. The two variants with one more degree of freedom than mirt's
#|   default overlap."
#| fig-alt: "Three panels, one per item set, plotting per-item rejection rate
#|   against sample size for three S-X2 variants. With mirt's df the rate is 6
#|   to 7.5 percent in every panel, above the dashed 5 percent line. With df + 1
#|   and with the variance fixed the two lines lie on top of each other at about
#|   5 percent."
dg_p |>
  filter(variant %in% c("V0", "V1", "V3")) |>
  group_by(set, variant, n) |>
  summarise(v = 100 * mean(p < .05, na.rm = TRUE), .groups = "drop") |>
  variant_plot("Items rejected at .05 (%)") +
  scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 2))
```

Adding the missing degree of freedom brings every cell to nominal: V1 rejects
4.8 to 5.3 percent of items against 6.0 to 7.5 under V0. V1 and V3 agree to
within 0.2 points in every cell, so estimating the latent variance costs nothing
and the df is the whole of the per-item excess. Under mirt's df, S-X2 exceeds
its df by 0.96 to 1.07 on average for dichotomous items, and with the variance
fixed the mean excess is within 0.07 of zero (@tbl-sx2-moments). The polytomous
set shows the same pattern from n = 500. At n = 150 its excess is 1.66 under V0
and 0.64 under V3, so something else adds to the statistic there.

Stricter collapsing on its own (V2) is worse than the default, because
`mincell = 5` merges so many score groups at n = 150 that the df falls to about
2, and a one-df shortfall weighs more at small df. With the df corrected (V2c)
it is close to nominal for dichotomous items, but at n = 150 it leaves too few
score groups to say much (@tbl-sx2-null-peritem).

### The far tail and the family-wise rate {#sec-sx2-null-fw}

BH over 20 items rejects on p-values near $.05 / 20$, so the family-wise rate
depends on the far tail of the p-value distribution, not on its 5 percent point.

```{r}
#| label: fig-sx2-null-fw
#| fig-width: 8
#| fig-height: 3.8
#| fig-cap: "Share of datasets with at least one item flagged by S-X2 with BH
#|   under a true Rasch model, 2000 datasets per sample size. The band covers two
#|   Monte Carlo SEs around 5 percent."
#| fig-alt: "Three panels plotting the family-wise false positive rate against
#|   sample size. In both dichotomous panels mirt's df gives 10.7 to 12.5
#|   percent, and the two corrected variants give about 7 to 8.5 percent, still
#|   above the band around 5 percent. In the polytomous panel the corrected
#|   variants lie within or near the band and mirt's df is about 6 to 8
#|   percent."
dg_p |>
  filter(variant %in% c("V0", "V1", "V3")) |>
  group_by(set, variant, n, rep) |>
  summarise(any = any(p.adjust(p, "BH") < .05, na.rm = TRUE), .groups = "drop") |>
  group_by(set, variant, n) |>
  summarise(v = 100 * mean(any), .groups = "drop") |>
  variant_plot("Datasets with any BH flag (%)",
               band = 100 * 2 * sqrt(.05 * .95 / 2000)) +
  scale_y_continuous(limits = c(0, 14), breaks = seq(0, 14, 2))
```

Correcting the df leaves part of the problem. Under V1, items reach p < .005 at
6.4 to 7.7 per 1000 for dichotomous items against a nominal 5, and BH flags at
least one item in 6.8 to 8.6 percent of datasets. That is down from 10.7 to
12.5 under V0 but still above 5. The polytomous set is close to nominal under
V1, at 4.3 to 6.5 percent. V2c removes the tail excess for dichotomous items
(3.4 to 5.3 per 1000 and 3.6 to 5.9 percent family-wise), which points to
sparse score groups as the source of what remains (@tbl-sx2-null-tail,
@tbl-sx2-null-fw). None of the rates rises steadily with sample size.

## S-X2 against conditional infit {#sec-sx2-comparison}

### Design {#sec-sx2-comp-design}

- **Items.** 20 dichotomous items, locations evenly spaced on [-2, 2].
- **Persons.** A bivariate normal, both dimensions SD 1.5, correlation $\rho$.
  Fitting items respond to the first dimension, misfitting items to the second.
- **Misfit.** k = 1 misfitting item (nearest 0 logits) or k = 3 (nearest 0, -1
  and -2). $\rho$ = 0.15 is strong multidimensionality, 0.50 and 0.75 are
  subtle.
- **Sample sizes.** 150, 500, 1000 and 2000. 500 replications in each null cell
  (k = 0) and 200 in each misfit cell. The Monte Carlo SE of a 5 percent
  family-wise rate is 1.0 point.

| Label | Method |
|---|---|
| Infit, Westfall-Young | conditional infit, `RMitemInfitCutoff()` with B = 400, Westfall-Young step-down, the easyRasch2 default |
| Infit, Holm | same bootstrap, Holm on the marginal p |
| S-X2, mirt df | chi-square on mirt's df, Holm |
| S-X2, df + 1 | chi-square on df + 1, Holm |
| S-X2, df + 1, BH | chi-square on df + 1, BH |
| S-X2, bootstrap | parametric bootstrap with B = 400 and a refit per draw, Westfall-Young |
| RMSEA, bootstrap | RMSEA.S_X2 from the same draws, Westfall-Young |
| RMSEA > .05 | a fixed threshold, as is sometimes recommended |

Every method is corrected for a family-wise error rate, with Westfall-Young or
Holm, except the BH variant. The design written before the run used BH for the
asymptotic S-X2 arms. BH controls the false discovery rate instead, and once the
misfitting items are rejected it loosens its threshold for the rest, so it flags
more fitting items than a family-wise method would. Holm was added after the run
from the stored p-values to make the comparison fair, and BH is kept in the
appendix tables to show the difference.

The S-X2 bootstrap draws datasets from the fitted Rasch model, refits each and
computes S-X2 for every item. The number of score groups changes between draws,
so the bootstrap statistic is $z = \Phi^{-1}(1 - p)$, with $p$ from the
chi-square on df + 1, which is on a common scale whatever the df. It goes
through the same Westfall-Young step-down that easyRasch2 uses for infit, so the
two arms differ only in the statistic. The design and a decision rule were
written before the run. The simulation is `scripts/sx2/sx2_comparison.R`, about
7.6 hours on 10 cores. Every replication returned results for every method and
no bootstrap draw was dropped.

```{r comp-tidy}
fs <- list.files("results/sx2/comparison", "\\.rds$", full.names = TRUE)
L <- lapply(fs, readRDS)
long <- do.call(rbind, lapply(L, function(x) do.call(rbind, lapply(seq_along(x$res), function(r) {
  y <- x$res[[r]]
  if (is.null(y$res)) return(NULL)
  cbind(k = x$cell$k, rho = x$cell$rho, n = x$cell$n, rep = r, y$res)
}))))

methods <- c("infit_wy", "infit_holm", "sx2_mirt", "sx2_df1", "sx2_df1_bh",
             "sx2_boot", "rmsea_boot", "rmsea_05")

adj <- long |>
  group_by(k, rho, n, rep) |>
  mutate(infit_wy = padj_infit, infit_holm = p.adjust(p_infit, "holm"),
         sx2_mirt = p.adjust(p_mirt, "holm"), sx2_df1 = p.adjust(p_df1, "holm"),
         sx2_df1_bh = p.adjust(p_df1, "BH"),
         sx2_boot = padj_sx2boot, rmsea_boot = padj_rmseaboot) |>
  ungroup()

flags <- adj |>
  mutate(across(c(infit_wy, infit_holm, sx2_mirt, sx2_df1, sx2_df1_bh, sx2_boot,
                  rmsea_boot),
                \(p) p < .05),
         rmsea_05 = rmsea > .05) |>
  pivot_longer(all_of(methods), names_to = "method", values_to = "flag") |>
  mutate(flag = coalesce(flag, FALSE),
         method = factor(method, levels = methods))

pal <- c(infit_wy = "#2a78d6", sx2_df1 = "#eb6834", sx2_boot = "#1baf7a",
         sx2_mirt = "#4a3aa7", rmsea_05 = "grey55")
shp <- c(infit_wy = 16, sx2_df1 = 17, sx2_boot = 15, sx2_mirt = 18,
         rmsea_05 = 4)
lty <- c(infit_wy = "solid", sx2_df1 = "solid", sx2_boot = "solid",
         sx2_mirt = "solid", rmsea_05 = "dashed")
lab <- c(infit_wy = "Infit, Westfall-Young", sx2_df1 = "S-X2, df + 1, Holm",
         sx2_boot = "S-X2, bootstrap", sx2_mirt = "S-X2, mirt df, Holm",
         rmsea_05 = "RMSEA > .05")

cell_lab <- function(d) d |>
  mutate(cell = factor(sprintf("k = %d, rho = %.2f", k, rho),
                       levels = c("k = 1, rho = 0.50", "k = 1, rho = 0.75",
                                  "k = 3, rho = 0.15", "k = 3, rho = 0.50",
                                  "k = 3, rho = 0.75")))

rate_plot <- function(d, ylab, ylim = c(0, 100)) {
  ggplot(d, aes(factor(n), v, colour = method, shape = method,
                linetype = method, group = method)) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 2.2) +
    scale_colour_manual(values = pal, labels = lab, name = NULL) +
    scale_shape_manual(values = shp, labels = lab, name = NULL) +
    scale_linetype_manual(values = lty, labels = lab, name = NULL) +
    scale_y_continuous(limits = ylim) +
    labs(x = "Sample size", y = ylab) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}

fig_methods <- c("infit_wy", "sx2_df1", "sx2_boot", "rmsea_05")
```

### Under the null {#sec-sx2-comp-null}

```{r}
#| label: fig-sx2-comp-null
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Share of datasets with at least one flagged item under a true
#|   Rasch model (k = 0, 500 datasets per n). The band covers two Monte Carlo
#|   SEs around 5 percent. RMSEA above .05 is left out, since it flags 99.2
#|   percent of datasets at n = 150."
#| fig-alt: "Line chart of family-wise false positive rate against sample size
#|   for four methods. Infit with Westfall-Young stays at about 4 percent at
#|   every n. The S-X2 bootstrap is about 6 percent at n = 150 and 500 and 4
#|   to 5 percent at n = 1000 and 2000, inside or at the edge of the band. S-X2
#|   with df + 1 is 6 to 9 percent and S-X2 with mirt's df 9 to 12 percent,
#|   both above the band at every n."
null_fw <- flags |>
  filter(k == 0) |>
  group_by(method, n, rep) |>
  summarise(any = any(flag), .groups = "drop") |>
  group_by(method, n) |>
  summarise(v = 100 * mean(any), .groups = "drop")

mcse <- 100 * 2 * sqrt(.05 * .95 / 500)
null_fw |>
  filter(method %in% c("infit_wy", "sx2_mirt", "sx2_df1", "sx2_boot")) |>
  rate_plot("Datasets with any flag (%)", ylim = c(0, 14)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 5 - mcse, ymax = 5 + mcse,
           fill = "grey85", alpha = 0.4) +
  geom_hline(yintercept = 5, linetype = "dashed", colour = "grey40")
```

Infit with Westfall-Young holds 4.2 to 4.4 percent at every n. The S-X2
bootstrap gives 6.2 and 6.6 percent at n = 150 and 500 and 4.8 and 3.8 percent
at n = 1000 and 2000. The first two are 1.2 and 1.6 Monte Carlo SEs above 5, so
it is at most slightly liberal at small n. Asymptotic S-X2 with Holm gives 5.8
to 8.6 percent with the df corrected, in line with @sec-sx2-null, and 8.6 to
12.0 with mirt's df. RMSEA above .05 flags almost every dataset at n = 150 and none at
n = 2000 (@tbl-sx2-comp-null).

RMSEA with a bootstrap cutoff behaves like the S-X2 bootstrap in every cell, as
expected, since both come from the same draws and RMSEA is close to a monotone
function of S-X2. Infit with Holm on the marginal bootstrap p is close to infit
with Westfall-Young and slightly more conservative. Both are left out of the
figures and kept in the appendix tables.

### Detection of the misfitting items {#sec-sx2-comp-detection}

```{r}
#| label: fig-sx2-detection
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Detection of the misfitting items, averaged over the misfitting
#|   items in each dataset. k is the number of misfitting items and rho the
#|   correlation between the two dimensions, lower meaning stronger misfit."
#| fig-alt: "Five panels, one per combination of number of misfitting items
#|   and dimension correlation, each plotting detection rate from 0 to 100
#|   percent against sample size. Detection rises with n in every panel.
#|   Infit with Westfall-Young is highest or tied in every panel. The S-X2
#|   bootstrap and S-X2 with df + 1 lag behind infit most in the two panels with
#|   rho = 0.75, catching up only where all methods reach 100 percent. RMSEA
#|   above .05, a grey dashed line, is high at n = 150 and falls with n in the
#|   rho = 0.75 panels."
flags |>
  filter(k > 0, misfit, method %in% fig_methods) |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(flag), .groups = "drop") |>
  cell_lab() |>
  rate_plot("Detection (%)") +
  facet_wrap(~ cell, nrow = 2)
```

Infit with Westfall-Young detects at least as much as the S-X2 bootstrap in
every cell, and clearly more wherever detection is not saturated. The gap is
largest for subtle misfit. At $\rho$ = 0.75 it is 80 against 47.5 percent with
one misfitting item at n = 500, and 58 against 32 percent with three at
n = 1000. The S-X2 bootstrap only catches up where both methods reach 100
percent (@tbl-sx2-detection).

The fixed RMSEA threshold detects more at n = 150 because it flags almost
everything there, and falls to 1.3 percent at $\rho$ = 0.75 and n = 2000,
because the RMSEA of a subtly misfitting item settles below .05 as n grows.
With three misfitting items at $\rho$ = 0.75, their median RMSEA falls from .047
at n = 150 to .031 at n = 2000. A
bootstrap cutoff for RMSEA recovers 68.7 percent in the same cell, but no more
than the S-X2 bootstrap.

### False flags on fitting items {#sec-sx2-comp-falseflags}

```{r}
#| label: fig-sx2-falseflags
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Share of datasets in which at least one fitting item is flagged.
#|   Panels as in @fig-sx2-detection."
#| fig-alt: "Five panels plotting the share of datasets with a falsely flagged
#|   fitting item against sample size. With three misfitting items, infit with
#|   Westfall-Young rises steeply with n, to 87 and 100 percent at n = 2000 for
#|   rho 0.50 and 0.15, while the S-X2 bootstrap stays below 10 percent except
#|   at rho = 0.15 and n = 2000, where it reaches about 48 percent. With one
#|   misfitting item all methods except RMSEA stay low. RMSEA above .05 is near
#|   100 percent at n = 150 in every panel."
flags |>
  filter(k > 0, !misfit, method %in% fig_methods) |>
  group_by(k, rho, n, method, rep) |>
  summarise(any = any(flag), .groups = "drop") |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(any), .groups = "drop") |>
  cell_lab() |>
  rate_plot("Datasets with a fitting item flagged (%)") +
  facet_wrap(~ cell, nrow = 2)
```

This is where S-X2 differs. With three misfitting items, infit flags fitting
items in a share of datasets that rises steeply with n: at $\rho$ = 0.50, 16
percent at n = 500, 39.5 at 1000 and 87 at 2000. The S-X2 bootstrap stays at 2
to 7.5 percent in the same cells. At $\rho$ = 0.15 infit flags fitting items in
97 percent of datasets at n = 1000 against 6.5 for the S-X2 bootstrap, and only
at n = 2000 does the S-X2 bootstrap also cascade, to 47.5 percent. Counted per
item at $\rho$ = 0.15 and n = 2000, infit flags 57.1 percent of the fitting
items and the S-X2 bootstrap 3.7 percent. With a single misfitting item the
cascade is milder for infit, up to 21 percent at $\rho$ = 0.50 and n = 2000,
against 5 percent (@tbl-sx2-falseflags).

Asymptotic S-X2 with the df corrected and Holm resists the cascade almost as
well as the bootstrap. With three misfitting items it flags fitting items in 2.5,
4.5 and 9 percent of datasets at $\rho$ = 0.50 and n = 500, 1000 and 2000,
against 2, 3.5 and 7.5 for the bootstrap, and its detection is as high or
slightly higher. With BH the same p-values flag fitting items in 7, 14.5 and 28
percent of datasets, which is the effect of the correction, not of the
statistic. Infit with Holm still cascades, to 72.5 percent at $\rho$ = 0.50 and
n = 2000, so the cascade under infit is not a property of Westfall-Young either.

A likely reason is what each statistic is sensitive to. The cascade leaves the
fitting items with small deviations in slope relative to the misspecified
model. Infit is close to a one-degree-of-freedom contrast in slope, so it picks
such deviations up, while S-X2 spreads them over 15 to 17 degrees of freedom at
n = 1000 and 2000 and has little power against them.

The fitting items are fitting by construction, since they respond only to the
first dimension. In the unidimensional model fitted to the whole set, the
misfitting items distort the person locations, so fitting items can show misfit
too [@johansson2025]. Flagging them is a false positive with respect to the
data generating model. Multiplicity control does not prevent it, since at large
n those items are far outside their null distribution.

### Recovering the right items {#sec-sx2-comp-perfect}

```{r}
#| label: fig-sx2-perfect
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Share of datasets in which every misfitting item is flagged and no
#|   fitting item is. Panels as in @fig-sx2-detection."
#| fig-alt: "Five panels plotting perfect recovery against sample size. With
#|   three misfitting items at rho 0.15 and 0.50, the S-X2 bootstrap climbs to
#|   about 93 percent at n = 1000 or 2000 while infit with Westfall-Young falls
#|   towards zero at n = 2000. At rho = 0.75 infit is higher than the S-X2
#|   bootstrap at every n. With one misfitting item both methods are high from
#|   n = 500, and the S-X2 bootstrap is higher at n = 2000 for rho = 0.50."
flags |>
  filter(k > 0, method %in% fig_methods) |>
  group_by(k, rho, n, method, rep) |>
  summarise(v = all(flag[misfit]) & !any(flag[!misfit]), .groups = "drop") |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(v), .groups = "drop") |>
  cell_lab() |>
  rate_plot("Perfect recovery (%)") +
  facet_wrap(~ cell, nrow = 2)
```

The two effects trade off. With three misfitting items the S-X2 bootstrap
recovers the right set far more often at n $\geq$ 1000 when misfit is strong or
moderate: 93.5 against 3 percent at $\rho$ = 0.15 and n = 1000, and 92.5 against
13 percent at $\rho$ = 0.50 and n = 2000. Infit is better at small n and for
subtle misfit, where the S-X2 bootstrap misses too many of the misfitting items:
41.5 against 26.5 percent at $\rho$ = 0.75 and n = 2000, and 47 against 30
percent at $\rho$ = 0.50 and n = 500 (@tbl-sx2-perfect).

### Detection by item location {#sec-sx2-comp-location}

```{r}
#| label: fig-sx2-location
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Detection of each of the three misfitting items by its location
#|   relative to the sample mean, for the two subtler conditions."
#| fig-alt: "Two rows of three panels, rows for rho 0.50 and 0.75, columns for
#|   misfitting items at about -2, -1 and 0 logits, each plotting detection
#|   against sample size for infit with Westfall-Young and the S-X2 bootstrap.
#|   Detection is lowest for the item at -2 logits in both rows. At rho = 0.75
#|   infit is well above the S-X2 bootstrap for every item, most of all for the
#|   two items nearer the centre."
flags |>
  filter(k == 3, misfit, rho > 0.15, method %in% c("infit_wy", "sx2_boot")) |>
  mutate(location = factor(sprintf("item at %.2f logits", location),
                           levels = sprintf("item at %.2f logits",
                                            sort(unique(location)))),
         rho = sprintf("rho = %.2f", rho)) |>
  group_by(rho, n, location, method) |>
  summarise(v = 100 * mean(flag), .groups = "drop") |>
  rate_plot("Detection (%)") +
  facet_grid(rho ~ location)
```

Both methods find it hardest to detect the item two logits below the sample
mean, and the S-X2 bootstrap loses most relative to infit on the two items
closer to the centre at $\rho$ = 0.75 (@tbl-sx2-location).

### At a matched null error rate {#sec-sx2-comp-matched}

The S-X2 bootstrap is slightly above 5 percent under the null at small n and
infit slightly below, which favours S-X2 in the comparisons above. In
@tbl-sx2-matched each method's adjusted p-value threshold is set so that its
family-wise rate in the null cell at the same n is 5 percent, and that
threshold is applied to the misfit cells. The thresholds come from the same
design, so they are an oracle.

Matching the null error rate moves detection by at most 3.3 points. False flags
move more where a threshold loosens, by up to 6.5 points for infit and 13.5 for
the S-X2 bootstrap, both at $\rho$ = 0.15, where the cascade is steepest. Infit
still detects more in every unsaturated cell, and the S-X2 bootstrap still flags
far fewer fitting items when three items misfit.

### Paired comparison {#sec-sx2-comp-paired}

Both methods were run on the same datasets, so their decisions can be compared
item by item. @tbl-sx2-paired counts misfitting items detected by one method
and not the other. Where the two disagree about a misfitting item, it is almost
always infit that detects it. The largest counts in the other direction, 15 to
17 out of 600, are in cells where infit alone detects 92 to 172.

## Interpretation {#sec-sx2-interpretation}

**1. mirt's S-X2 is liberal for Rasch models.** Subtracting the latent variance
from every item's df makes S-X2 reject 6 to 7.5 percent of items at a nominal 5
under a true Rasch or partial credit model. Adding the df back, or fixing the
variance, gives 4.8 to 5.3. Whether the adjustment is right for other models is
not tested here.

**2. Correcting the df is not enough for a family of dichotomous items.** A
smaller excess in the far tail, traced to sparse score groups, keeps the
family-wise rate at 6.8 to 8.6 percent for 20 dichotomous items. A parametric
bootstrap removes it, apart from a slight excess at small n.

**3. Infit with Westfall-Young is calibrated and detects more.** It holds 4.2 to
4.4 percent family-wise under the null and detects at least as much as the S-X2
bootstrap and the df-corrected S-X2 in every cell, at nominal and at matched
error rates. For subtle misfit and moderate n the difference is large.

**4. S-X2 resists the cascade.** When three items misfit at n $\geq$ 1000, it
rarely flags fitting items, while infit flags them in most datasets. Its
advantage in recovering the right set of items is large where misfit is strong
or moderate and n is large. The resistance belongs to the S-X2 statistic, since
asymptotic S-X2 with Holm resists about as well as the bootstrap. The bootstrap's
contribution is calibration under the null, where asymptotic S-X2 with Holm is
still at 5.8 to 8.6 percent.

**5. RMSEA adds nothing beyond S-X2.** A fixed threshold of .05 is unusable in
both directions. A bootstrap cutoff makes RMSEA usable, but it then behaves like
the S-X2 bootstrap.

**Against the decision rule.** The rule set before the run was that S-X2 is
worth adding if its bootstrap version holds 5 percent under the null and, in a
cell with $\rho$ = 0.50 or 0.75, detects misfit infit misses or flags clearly
fewer fitting items, without a comparable loss elsewhere. The first two parts
are met. At $\rho$ = 0.50 with three misfitting items, the S-X2 bootstrap flags
fitting items in 3.5 against 39.5 percent of datasets at n = 1000 and 7.5
against 87 at n = 2000, with detection of 93.8 against 98 and 100 against 100.
The third part is a judgement, since at $\rho$ = 0.75 the S-X2 bootstrap detects
much less than infit at every n up to 1000.

So S-X2 is neither a replacement for infit nor a better default. Its use is as
a second opinion at large n: when infit flags many items, S-X2 separates the
items that misfit from those that are flagged because others do. The bootstrap
version holds the null and costs a refit per draw, about 25 to 40 seconds per
analysis at B = 400 on one core for 20 items. Asymptotic S-X2 with the df
corrected and Holm gives nearly the same second opinion at no cost, with a
family-wise error rate of 6 to 9 percent instead of 5.

## Scope and limits {#sec-sx2-scope}

- The comparison uses dichotomous items only. The null study found S-X2 close
  to nominal for nine polytomous items after the df correction, so a polytomous
  comparison is the obvious next step.
- One misfit mechanism, a second dimension. Other mechanisms, such as
  discrimination different from the Rasch model, may favour either method.
- S-X2 is computed by mirt under MML with a normal latent distribution, which
  matches the data generation here and so favours it slightly.
- The bootstrap uses df + 1 only to put S-X2 on a common scale across draws. Its
  reference does not depend on the df otherwise.

## Appendix: result tables {#sec-sx2-appendix}

```{r appendix-setup}
#| include: false
# kable with rho printed to two decimals, so digits = 1 does not round it
kb <- function(d, ...) {
  d <- mutate(d, across(any_of("rho"), \(x) sprintf("%.2f", x)))
  knitr::kable(d, ...)
}
wide_n <- function(d, value) d |>
  pivot_wider(names_from = n, values_from = all_of(value), names_prefix = "n = ")
main <- c("infit_wy", "infit_holm", "sx2_df1", "sx2_df1_bh", "sx2_boot",
          "rmsea_05")
```

### S-X2 under a true Rasch model

```{r}
#| label: tbl-sx2-moments
#| tbl-cap: "Mean S-X2 minus its df, with mirt's df (V0) and with the variance
#|   fixed (V3), and the variance of S-X2 relative to twice df + 1 under V0."
dg_long |>
  group_by(set, n) |>
  summarise(`S - df, V0` = mean(S_v0 - df_v0, na.rm = TRUE),
            `S - df, V3` = mean(S_v3 - df_v3, na.rm = TRUE),
            `var / 2(df + 1), V0` = var(S_v0 - df_v0, na.rm = TRUE) /
              (2 * mean(df_v0 + 1, na.rm = TRUE)),
            `mean df, V0` = mean(df_v0, na.rm = TRUE), .groups = "drop") |>
  kb(digits = 2)
```

```{r}
#| label: tbl-sx2-null-peritem
#| tbl-cap: "Per-item rejection at .05 (%), all five variants."
dg_p |>
  group_by(set, variant, n) |>
  summarise(rate = 100 * mean(p < .05, na.rm = TRUE), .groups = "drop") |>
  wide_n("rate") |>
  kb(digits = 2)
```

```{r}
#| label: tbl-sx2-null-tail
#| tbl-cap: "Items with p < .005, per 1000 (nominal 5)."
dg_p |>
  group_by(set, variant, n) |>
  summarise(v = 1000 * mean(p < .005, na.rm = TRUE), .groups = "drop") |>
  wide_n("v") |>
  kb(digits = 1)
```

```{r}
#| label: tbl-sx2-null-fw
#| tbl-cap: "Datasets with at least one BH-flagged item (%)."
dg_p |>
  group_by(set, variant, n, rep) |>
  summarise(any = any(p.adjust(p, "BH") < .05, na.rm = TRUE), .groups = "drop") |>
  group_by(set, variant, n) |>
  summarise(v = 100 * mean(any), .groups = "drop") |>
  wide_n("v") |>
  kb(digits = 2)
```

### S-X2 against conditional infit

```{r}
#| label: tbl-sx2-comp-null
#| tbl-cap: "Datasets with at least one flagged item under the null (%), k = 0."
null_fw |>
  wide_n("v") |>
  kb(digits = 1)
```

```{r}
#| label: tbl-sx2-detection
#| tbl-cap: "Detection, mean over the misfitting items (%)."
flags |>
  filter(k > 0, misfit, method %in% main) |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(flag), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = v) |>
  kb(digits = 1)
```

```{r}
#| label: tbl-sx2-falseflags
#| tbl-cap: "Datasets with at least one fitting item flagged (%)."
flags |>
  filter(k > 0, !misfit, method %in% main) |>
  group_by(k, rho, n, method, rep) |>
  summarise(any = any(flag), .groups = "drop") |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(any), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = v) |>
  kb(digits = 1)
```

```{r}
#| label: tbl-sx2-perfect
#| tbl-cap: "Perfect recovery: every misfitting item flagged and no fitting item (%)."
flags |>
  filter(k > 0, method %in% main) |>
  group_by(k, rho, n, method, rep) |>
  summarise(ok = all(flag[misfit]) & !any(flag[!misfit]), .groups = "drop") |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(ok), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = v) |>
  kb(digits = 1)
```

```{r}
#| label: tbl-sx2-location
#| tbl-cap: "Detection by item location, k = 3 (%)."
flags |>
  filter(k == 3, misfit, rho > 0.15, method %in% c("infit_wy", "sx2_boot")) |>
  mutate(location = round(location, 2)) |>
  group_by(rho, n, location, method) |>
  summarise(v = 100 * mean(flag), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = v) |>
  kb(digits = 1)
```

```{r}
#| label: tbl-sx2-matched
#| tbl-cap: "Detection and false flags with each method's threshold set to a 5
#|   percent family-wise rate in the null cell at the same n (%)."
thr <- adj |>
  filter(k == 0) |>
  select(n, rep, infit_wy, sx2_df1, sx2_boot) |>
  pivot_longer(c(infit_wy, sx2_df1, sx2_boot), names_to = "method",
               values_to = "padj") |>
  group_by(method, n, rep) |>
  summarise(mn = min(padj, na.rm = TRUE), .groups = "drop") |>
  group_by(method, n) |>
  summarise(c = quantile(mn, .05, type = 1), .groups = "drop")

adj |>
  filter(k > 0) |>
  select(k, rho, n, rep, misfit, infit_wy, sx2_df1, sx2_boot) |>
  pivot_longer(c(infit_wy, sx2_df1, sx2_boot), names_to = "method",
               values_to = "padj") |>
  left_join(thr, by = c("method", "n")) |>
  mutate(flag = coalesce(padj <= c, FALSE)) |>
  group_by(k, rho, n, method, rep) |>
  summarise(det = mean(flag[misfit]), ff = any(flag[!misfit]), .groups = "drop") |>
  group_by(k, rho, n, method) |>
  summarise(det = 100 * mean(det), ff = 100 * mean(ff), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = c(det, ff)) |>
  select(k, rho, n, det_infit_wy, det_sx2_boot, ff_infit_wy, ff_sx2_boot) |>
  kb(digits = 1, col.names = c("k", "rho", "n", "Detection, infit",
                               "Detection, S-X2 boot", "False flags, infit",
                               "False flags, S-X2 boot"))
```

```{r}
#| label: tbl-sx2-paired
#| tbl-cap: "Misfitting items detected by one method and not the other, infit
#|   with Westfall-Young against the S-X2 bootstrap, same datasets."
flags |>
  filter(k > 0, misfit, method %in% c("infit_wy", "sx2_boot")) |>
  select(k, rho, n, rep, item, method, flag) |>
  pivot_wider(names_from = method, values_from = flag) |>
  group_by(k, rho, n) |>
  summarise(`infit only` = sum(infit_wy & !sx2_boot),
            `S-X2 boot only` = sum(!infit_wy & sx2_boot),
            both = sum(infit_wy & sx2_boot), .groups = "drop") |>
  kb()
```

## Revision history {.unnumbered}

- 2026-10-05: first version, combining `easyRasch2/dev/sx2_null_diagnostic.qmd`
  and `easyRasch2/dev/sx2_comparison_v2.qmd`. Replaces an unpublished July 2026
  comparison whose design had problems, including infit without multiplicity
  correction and replications drawn from a single simulated dataset.
- 2026-10-05: asymptotic S-X2 and the infit comparison arm now use Holm, a
  family-wise correction, instead of BH. The first version credited the S-X2
  bootstrap with the resistance to the cascade, because asymptotic S-X2 with BH
  cascaded more. With Holm it does not, so the resistance belongs to S-X2
  itself. Interpretation 4 and the summary are corrected.
